DRAFT \\ From physical principles to classical and quantum \\ Hamiltonian and Lagrangian particle mechanics

### 2017-01-08 2015CompleteDerivation/completeDerivation.tex
Using theorem env provided by journal
Reviewed intro/conclusions with Christine

%\documentclass[aps,pra,10pt,twocolumn,floatfix,nofootinbib]{revtex4-1}
\documentclass[smallextended,nospthms]{svjour3}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{dutchcal}
\usepackage{braket}
\usepackage{enumitem}

@@

@@

\newtheorem{assump}{Assumption}
\renewcommand*{\theassump}{\Roman{assump}}
\newtheorem{thrm}[equation]{Theorem}

\theoremstyle{definition}
\newtheorem{prop}[equation]{Proposition}

\newenvironment{rationale}{\emph{Rationale}.}{\qed}
\newenvironment{justification}{\emph{Justification}.}{\qed}
\renewenvironment{proof}{\emph{Proof}.}{\qed}

@@

\institute{G. Carcassi \at
	University of Michigan, Ann Arbor, MI 48109 \\
%	Tel.: +123-45-678910\\
%	Fax: +123-45-678910\\
	\email{carcassi@umich.edu}           %  \\
	%             \emph{Present address:} of F. Author  %  if needed
}

\titlerunning{From physical principles to classical and quantum particle mechanics}

@@

While some branches of physics (e.g. Newtonian mechanics(ref?), thermodynamics(ref?), special relativity(ref?)) are derived from physical principles (e.g. Newton's laws, laws of thermodynamics, invariance of the speed of light), others start by simply setting a mathematical framework. It is the case for classical Lagrangian mechanics(ref?), where we start by assuming the existence of a Lagrangian function whose path integral is minimized during the motion. It is the case for classical Hamiltonian mechanics(ref?), where we start by assuming the state is described by conjugate pairs $(q^i, p_i)$ and evolves according to Hamilton's equations. It is the case for quantum mechanics(ref?), where the state space is a complex inner product vector space and the state evolves according to Schroedinger's equations. That is, at the very core of some of our physical theories we have mathematical definitions instead of physical ideas.

If we ask ``when can a particular physical system be suitably described by classical Hamiltonian mechanics?" the only answer we have is ``when it fits the mathematical framework". While many seem to be content with this answer (e.g. ``you have to start from somewhere, it may as well be the math which at least is precise"), it begs the question: it does not tell us how and what physical properties of the system under study are being captured.

This is a problem since mathematical structures are not enough to characterize the physical system being described. For example, the same mathematical framework for linear circuits can be used to describe electric, hydraulic, thermal or mechanical systems.\cite{Roland,Barron,Borutzky} The mathematical treatment is the same because it captures aspects of the description that are common to all cases. But for the very same reason, we cannot infer what system is being described if all we are given is the mathematical model. In the jargon of linguistic/computer science/phylosophy\cite{Freiding,Friedman}: math captures the syntax (the relationships between the objects) but does not capture the semantic (the meaning of each object).

We started this work with the unpretentious goal of exploring simple questions such as the following. What do the mathematical symbols used in classical/quantum Hamiltonian represent? What is the physical meaning attached to it? Why is the state of a quantum particle described by a wave function (i.e. a vector in a complex inner product space), while the state of a classical particle is described by the pairs $(q^i, p_i)$ (i.e. a point in a cotangent bundle)? The idea was to create a math-to-physics dictionary so that each mathematical concept was clearly mapped to a crisp physical concept. But to make sure that such dictionary was complete, we realized we had to make sure that the whole relevant mathematical framework could be recovered: the translation needs to work both ways.

This redefined our mission to re-deriving as much as possible of the mathematical structure by starting from the smallest set of physical assumptions. It also forced us to push further and further back toward the basic definitions. For example, if we truly wanted to justify why the state space of Hamiltonian mechanics is charted by conjugate variables $q^i$ and $p_i$, we had to justify why it was a manifold as well, which in turn meant we needed to justify why it was a topological space. But when we did find physical concepts that mapped to these more basic mathematical definitions, we realized that this link was valid for a far wider range of cases: the concepts were more general than we anticipated. For example, we realized that the mathematical notion of a topology maps incredibly well to the idea of physical distinguishability, but states of Hamiltonian system are not the only thing that are physically distinguishable.

Each new link often shrank the complexity of the arguments while expanding the scope of the work in a way that was not at all anticipated. The same link that allowed to justify Hamiltonian dynamics also gave us relativistic motion on the extended phase space. The same link that allowed to justify Lagrangian mechanics also constrained the motion to that of free particles under scalar/vector potential forces. We also realized that sometimes different mathematical concepts could be justified from the same physical starting point. That is: by putting physics first we could achieve a sort of unification of ideas that was not evident in the math alone.

The unexpected success of the approach lead us to think that, maybe, the picture we were developing was not simply ``a nicer way to think about things", but instead a better framework to understand and link different parts of basic science and mathematics. After many years of work and different incremental iterations, we think the arguments are well developed enough to report the progress.

The goal of this paper is to show that it is possible to derive classical/quantum Hamiltonian/Lagrangian mechanics starting from few physical assumptions. We do \emph{not} claim that we have perfectly achieved this result, with no room from improvement. We do claim that the whole architectural framework is solid enough to be convinced that this is indeed possible, and that it is a worthwhile endeavor as it allows to make intuitive connections between different areas of physics and mathematics. That is: it \emph{can} and \emph{should} be done.

While trying to re-derive classical, quantum and (to some extent) relativistic mechanics may seem too broad a scope, it is necessary to show that the concepts we develop are truly general, that they can be applied to many different cases. If we could derive classical Hamiltonian mechanics on assumptions that are inapplicable to the quantum case, what good would it do? Quantum system are Hamiltonian systems, so clearly we wouldn't have understood what makes a system Hamiltonian. The ability to apply the same physical ideas far and wide is what gives us confidence that these are truly general concept that capture the physical aspects we are describing. The breath itself is what makes the work interesting and more than a mere curiosity.

As a brief overview, there are four assumptions we will consider. Assuming deterministic and reversible evolution will give us a dynamical system: states and a law of evolution. Additionally assuming infinitesimal reducibility, studying the whole system is equivalent to studying its infinitesimal parts, gives us Hamiltonian mechanics. Additionally assuming kinematic equivalence, trajectories in space tell us everything about the state of the system, gives us Lagrangian mechanics and constrains the motion to massive particles under scalar/vector potential forces. Relativistic motion arises when studying time dependent laws of evolution with no further assumption. Alternatively, as the second assumption we can take irreducibility, studying the whole systems tells us nothing about its parts, which leads to quantum (Hamiltonian) mechanics.

As a more detailed overview: in section \ref{sec:fundamental_model} we will describe in general what it means to study a physical system and how the assumption of determinism and reversibility plays into it. We will develop an abstract conceptual model that we will use when discussing the later assumptions.

In section \ref{sec:dynamical_systems} we will show how the physical concept of physical distinguishability leads to the mathematical concept of topological spaces. Deterministic and reversible evolution is an invertible continuous map (i.e. a homeomorphism) as it needs to preserve what is physically distinguishable.

In section \ref{sec:reducibility} we will introduce the notion of a classical material, one for which we assume infinitesimal reducibility. The state of an arbitrary amount of material is a distribution over the states of the infinitesimal parts, the classical particles. As such distribution must be invariant under the arbitrary choice of units, we'll see that for each variable $q^i$ that defines a unit there is a corresponding variable $k^i$ defined on the inverse of such unit. Classical particle states can be identified by conjugate pairs of state variable $(q^i, p_i)$ (i.e. a point in the cotangent bundle $T^*\mathcal{Q}$) and the number of possible initial conditions for each pair is given by $dq^i \wedge dp_i$ (i.e. the canonical symplectic form $\omega$). During evolution, the distribution (and marginal distributions) must be mapped point by point which leads to Hamiltonian mechanics (i.e. a symplectomorphism on $T^*\mathcal{Q}$).

In section \ref{sec:relativity} we will show that a relativistic version of Hamilton's equations arises naturally when we consider time dependent laws of evolution for distributions of classical material. This also gives the notion of deterministic and reversible evolution a characterization that is independent of any transformation that mixes time and state variables $(q^i, p_i)$. It will also give us a classical equivalent for anti-particles states.

In section \ref{sec:Lagrangian} we will introduce the assumption of kinematic equivalence, studying the motion in space is equivalent to studying the evolution of the state. Under that assumption there is a link between the state variables $(q^i, p_i)$ and the kinematic variable $(x^i, u^i)$ (i.e. position and velocity). This link will allow us to express the number of possible initial conditions for a d.o.f. in terms of $dx^i$ and $du^i$, leading to the metric tensor $g$. Such expression also allows to constrain the Hamiltonian to the one for massive particles under scalar potential forces and an equivalent Lagrangian formulation.

In section \ref{sec:quantum_prelude} we will introduce some ideas in classical mechanics to more gently transition over quantum mechanics. We will see how classical materials (i.e. distributions over phase space) already have features that are typically associated with quantum systems and we will show how to construct a classical uncertainty principle that helps to build intuition used in the later section.

In section \ref{sec:irreducibility} we will introduce the notion of an irreducible material, one for which studying the whole distribution does not tell us anything about the motion of its infinitesimal parts, the fragments. A unit amount will correspond to a quantum particle. As this means that the states of such material are invariant under permutations of the configurations for the fragments, we will show that the state space of an irreducible material is a complex vector space. These states can still be compared to each other in a way that gives the state space an inner product. Such comparisons will be preserved by deterministic and reversible evolution which leads to Schroedinger equation (i.e. a unitary transformation).

In section \ref{sec:discussion} we will discuss of how this work could be extended in different directions.

@@

We claim that $\mathsf{T}$ is second countable. Consider the set of outcomes of a process. As each can provide only a finite amount of information (e.g. is identified by a finite number of digits), the set of outcomes is countable. This remains true when combining processes: the number of processes we can combine is finite as is the number of times we can prepare the same state and the set of intersections obtained from the countable outcomes of a finite number of processes is still countable. Consider a basis for the topology. Suppose it is of cardinality greater than countable. Then the basis would be able to distinguish among states in a way that no physical process ever could. The topology could distinguish among states that are not physically distinguishable. The basis for the topology cannot be of cardinality greater than countable. By contradiction, $\mathcal{S}$ must be second countable.

@@

	We claim that for any fragment variable $f$ there exists a corresponding self-adjoint linear operator $F_{[q^i]} : \Psi(\mathcal{Q}, dq^i) \rightarrow C(\mathcal{Q}, \mathbb{C})$ such that $\int \psi^\dagger(q^i) F_{[q^i]}( \psi(q^i)) d^nq$ is, if it exists, the expectation of $f$ multiplied by the amount of material. Let $f$ be a fragment variable. There exists a corresponding self-adjoint linear operator $F_{[\hat{q}^i]} : \Psi(\hat{\mathcal{Q}}, d\hat{q}^i) \rightarrow C(\hat{\mathcal{Q}}, \mathbb{C})$ where $\hat{\mathcal{Q}}$ provides a maximal set of fragment variables. As both $\hat{\mathcal{Q}}$ and $\mathcal{Q}$ provide a maximal set, we can define $\Upsilon^{q^i}_{\hat{q}^i} \equiv \Upsilon_{\hat{q}^i} \circ \Upsilon_{q^i}^{-1}$ which transforms a wave function over $\mathcal{Q}$ into a wave function over $\hat{\mathcal{Q}}$. $\Upsilon^{q^i}_{\hat{q}^i}$ is an isomorphism as $\Upsilon_{\hat{q}^i}$ and $\Upsilon_{q^i}$ are isomorphisms. $\Upsilon^{q^i}_{\hat{q}^i}$ is unitary as $\Upsilon^{\hat{q}^i}_{q^i} \circ \Upsilon^{q^i}_{\hat{q}^i} = (\Upsilon^{q^i}_{\hat{q}^i})^{-1} \circ \Upsilon^{q^i}_{\hat{q}^i} = I$. We can extend $\Upsilon^{q^i}_{\hat{q}^i}$ on all continuous functions and define $F_{[q^i]} \equiv \Upsilon^{\hat{q}^i}_{q^i} \circ F_{[\hat{q}^i]} \circ \Upsilon^{q^i}_{\hat{q}^i}$. $F_{[q^i]}$ is a self-adjoint linear operator. $\int \psi^\dagger(q^i) F_{[q^i]}( \psi(q^i)) d^nq = \int \psi^\dagger(\hat{q}^i) F_{[\hat{q}^i]}( \psi(\hat{q}^i)) d^n \hat{q}$ is, if it exists, the expectation of $f$ multiplied by the amount of material.

@@

The state space for classical particles is the one that allows us to express invariant distributions: that is a requirement that comes from the whole. The fragments of quantum particles do not even have a well defined description: the whole tells us nothing about their configuration. In retrospect, this is not be surprising. Nobody is going to hand us down the laws of infinitesimal parts: we induce them by manipulating finite systems. Therefore what the whole allows us to infer about its part is physically fundamental.

The third notable aspect is that the assumptions have more to do with our ability to create a reliable description rather than essential features of the system itself. A system is not deterministic per se: it is a particular choice for its description (i.e. state) that may undergo deterministic evolution in some cases. In the same vein, a material is not homogeneous per se. Or infinitesimally reducible. In this context, the state, its properties and the law of evolution are the description of the system under our idealization, not some intrinsic feature of the system.

The fourth intriguing aspect is that, when framing classical and quantum mechanics in this way, there is a progression. First we assume a homogeneous infinitesimally reducible material. Then we assume an homogeneous irreducible material. Connecting the dots, the next assumption would be a inhomogeneous irreducible material: where the material is not in constant equilibrium and its particles do not all look the same. This different starting point may have potential connections to some of the problems we mentioned in the rationales for the assumptions and to other interesting physics questions in general.

As there would be different particles, they could be seen as different stable states of the same inhomogeneous material. In particular, since $\sigma_x \sigma_v \geq \frac{\hbar}{2m}$, particles in different generations (i.e. electron vs muon) could be similar stable configurations where the fragments of the material are more "packed" in terms of states labeled by the kinematic variables.

As we are not always at equilibrium, the state of the system will be in general a statistical mixture of distributions of material. The evolution may also take pure states (i.e. a mixed state with a single state with $100\%$ probability) and evolve them into mixed states, thus giving an account of what happens during a measurement or a decay.

The law of evolution may still be deterministic and reversible, in the sense that given a particular mixed state we can predict/reconstruct future/past mixed state, but it will not be time symmetric.  We can then tell that time flows in the direction where information is exchanged, where entropy increases. This would give us a time arrow.

These ideas are appealing because they seem to fit well in the narrative, but at this point they are just speculation. The only thing we can say for now is that the state space under this new assumption would be a vector space on a field that is neither $\mathbb{R}$ or $\mathbb{C}$. But it is not unreasonable that these will inspire new work in the future.

In the foreseeable future, however, we will work on the more mundane task of extending this work to field theories, for which a more concrete path is available. The state space for a quantized material, already introduced here, should be captured by the Fock space. The idea of a non-minimal material will force us to consider vector and tensor quantities at each point in the distribution, whose state should be captured by spin systems. Assuming that the vector potential forces themselves are dependent on the state of a material should hopefully give us the Hamiltonian for the fields. Whether this can be done will be a matter of working through the details, which, we are sure, are likely to bring other unexpected surprises.

@@

We have seen that much of fundamental physics can be derived from few simple physical assumption (and lots of hard work). In doing so we obtained a more unified picture of the different branches of fundamental physics centered around the idea of deterministic and reversible evolution of distributions of homogeneous materials. Though the work is extensive we can summarize a few important conclusions with the following points:

@@

\item Deterministic and reversible evolution is a one-to-one map that preserves the nature of the system (e.g. physical distinguishability, state count and decomposability). Therefore it is an isomorphism on the mathematical structure defined by the state space (e.g. homeomorphism, canonical transformation, unitary transformation), not just a state-to-state bijection (i.e. isomorphism of sets).

\item Classical Hamiltonian mechanics describes the evolution of a homogeneous material whose state is equivalent to the states of all its infinitesimal parts, the classical particles. The state space for classical particles is a symplectic manifold $(T^*\mathcal{Q}, \omega)$ as this allows to define coordinate invariant density distributions; deterministic and reversible evolution for such state space is a symplectomorphism (i.e. a canonical transformation).

@@

\item Quantum (Hamiltonian) mechanics describes the evolution of a homogeneous material for which the state can only be reduced up to finite parts, the quantum particles, whose internal motion is undetermined. The state space for quantum particles is a complex inner product space $(\Psi, \langle \cdot , \cdot \rangle)$; deterministic and reversible evolution for such state space is a unitary transformation.
\end{itemize}

While we are sure that some mathematical details can be improved, we feel that the overall ``conceptual castle" is solid. The starting concepts are simple, the conceptual mapping of those concepts into math is often straight forward and the intuitive ideas that emerge help to connect different branches of math and physics. This is unlikely a matter of simple coincidence.

While the scope of this work is limited to re-founding and re-understanding physical theories already experimentally established, it would not be unexpected that a better grasp of old theories may lead to insights and ideas for new ones.

@@

\bibitem{Freiding} Freidin, R., Lasnik, H.: Syntax critical concepts in linguistics. Routledge (2006)

>>>>

%\documentclass[aps,pra,10pt,twocolumn,floatfix,nofootinbib]{revtex4-1}
\documentclass[smallextended]{svjour3}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{amsfonts}
\usepackage{dutchcal}
\usepackage{braket}
\usepackage{enumitem}

@@

%----

% Theorem definitions using amsthm

%\usepackage{amsthm}

@@

%\newtheorem{assump}{Assumption}
%\renewcommand*{\theassump}{\Roman{assump}}
%\newtheorem{thrm}[equation]{Theorem}

%\theoremstyle{definition}
%\newtheorem{prop}[equation]{Proposition}

%\newenvironment{rationale}{\emph{Rationale}.}{\qed}
%\newenvironment{justification}{\emph{Justification}.}{\qed}
%\renewenvironment{proof}{\emph{Proof}.}{\qed}

%----

\spnewtheorem{assump}{Assumption}{\bf}{\it}
\renewcommand{\theassump}{\Roman{assump}}
\spnewtheorem{prop}[equation]{Proposition}{\bf}{\rm}
\spnewtheorem{thrm}[equation]{Theorem}{\bf}{\it}

\newenvironment{rationale}{\emph{Rationale}.}{\hfill\(\qed\)}
\newenvironment{justification}{\emph{Justification}.}{\hfill\(\qed\)}
\renewenvironment{proof}{\emph{Proof}.}{\hfill\(\qed\)}

@@

\institute{G. Carcassi \at
	University of Michigan, Ann Arbor, MI 48109 \\
	\email{carcassi@umich.edu}           %  \\
}

% As the page margins are so large, the full title does not fit
\titlerunning{From physical principles to classical and quantum particle mechanics}

@@

While some branches of physics (e.g. Newtonian mechanics, thermodynamics, special relativity) are derived from physical principles (e.g. Newton's laws, laws of thermodynamics, invariance of the speed of light), others start by simply setting a mathematical framework. This is the case for classical Lagrangian mechanics, where we start by assuming the existence of a Lagrangian function whose path integral is minimized during the motion, for classical Hamiltonian mechanics, where we start by assuming the state is described by conjugate pairs $(q^i, p_i)$ and evolves according to Hamilton's equations, and for quantum mechanics, where the state space is a complex inner product space and the state evolves according to Schroedinger's equation. That is, at the very core of some of our physical theories we have mathematical definitions instead of physical ideas.

If we ask ``when can a particular physical system be suitably described by classical Hamiltonian mechanics?" the only rigorous answer we have is ``when it fits the mathematical framework". While many seem to be content with this answer (e.g. ``you have to start from somewhere, it may as well be the math which at least is precise"), it begs the question: it does not tell us how and what physical properties of the system under study are being captured.

This is a problem since mathematical structures are not enough to characterize the physical system being described. For example, the same mathematical framework for linear circuits can be used to describe electric, hydraulic, thermal or mechanical systems~\cite{Roland,Barron,Borutzky}. The mathematical treatment is the same because it captures aspects of the description that are common to all cases. But for the very same reason, we cannot infer what system is being described if all we are given is the mathematical model. In the jargon of linguistics/computer science/philosophy~\cite{Freidin,Friedman}: math captures the syntax (the relationships between the objects) but does not capture the semantics (the meaning of each object).

We started this work with the unpretentious goal of exploring simple questions such as the following. What do the mathematical symbols used in classical/quantum Hamiltonian mechanics represent? What is the physical meaning attached to them? Why is the state of a quantum particle described by a wave function (i.e. a vector in a complex inner product space), while the state of a classical particle is described by position and conjugate momentum (i.e. a point in a cotangent bundle)? The idea was to create a math-to-physics dictionary so that each mathematical concept was clearly mapped to a crisp physical concept. But to make sure that such a dictionary was complete, we realized we had to make sure that the whole relevant mathematical framework could be recovered: the translation needs to work both ways.

This redefined our mission to rederiving as much as possible of the mathematical structure by starting from the smallest set of physical assumptions. It also forced us to push further and further back toward the basic definitions. For example, if we truly wanted to justify why the state space of Hamiltonian mechanics is charted by conjugate variables $q^i$ and $p_i$, we had to justify why it was a manifold as well, which in turn meant we needed to justify why it was a topological space. But when we did find physical concepts that mapped to these more basic mathematical definitions, we realized that this link was valid for a far wider range of cases: the concepts were more general than we anticipated. For example, we realized that the mathematical notion of a topology maps incredibly well to the idea of physical distinguishability, but states of Hamiltonian system are not the only things that are physically distinguishable.

Each new link often shrank the complexity of the arguments while expanding the scope of the work in a way that was not at all anticipated. The same link that allowed us to justify Hamiltonian dynamics also gave us relativistic motion on the extended phase space. The same link that allowed us to justify Lagrangian mechanics also constrained the motion to that of massive particles under scalar/vector potential forces. We also realized that sometimes different mathematical concepts could be justified from the same physical starting point. That is: by putting physics first we could achieve a sort of unification of ideas that was not evident in the math alone.

The unexpected success of the approach led us to think that, maybe, the picture we were developing was not simply ``a nicer way to think about things", but instead a better framework to understand and link different parts of basic science and mathematics. After many years of work and different incremental iterations, we think the arguments are well developed enough to report the progress.

The goal of this paper is to show that it is possible to derive classical/quantum Hamiltonian/Lagrangian particle mechanics starting from few physical assumptions. We do \emph{not} claim that we have perfectly achieved this result, with no room for improvement. We do claim that the whole architectural framework is solid enough to be convinced that this is indeed possible, and that it is a worthwhile endeavor as it allows us to make intuitive connections between different areas of physics and mathematics. That is: it \emph{can} and \emph{should} be done.

While trying to rederive classical, quantum and (to some extent) relativistic mechanics may seem too broad a scope, it is necessary to show that the concepts we develop are truly general, that they can be applied to many different cases. If we could derive classical Hamiltonian mechanics on assumptions that are inapplicable to the quantum case, what good would it do? Quantum systems are Hamiltonian systems, so clearly we wouldn't have understood what makes a system Hamiltonian. The ability to apply the same physical ideas far and wide is what gives us confidence that these are truly general concepts that capture the physical aspects we are describing. The breadth itself is what makes the work interesting and more than a mere curiosity.

As a brief overview, there are four assumptions we will consider. Assuming deterministic and reversible evolution gives us a dynamical system: states and a law of evolution. Additionally assuming infinitesimal reducibility, studying the whole system is equivalent to studying its infinitesimal parts, gives us Hamiltonian mechanics. Further assuming kinematic equivalence, trajectories in space tell us everything about the state of the system, gives us Lagrangian mechanics and constrains the motion to massive particles under scalar/vector potential forces. Relativistic motion arises when studying time dependent laws of evolution with no further assumptions. Alternatively, as the second assumption we can instead take irreducibility, studying the whole system tells us nothing about its parts, which leads to quantum (Hamiltonian) mechanics.

As a more detailed overview: in section \ref{sec:fundamental_model} we describe in general what it means to study a physical system and how the assumption of determinism and reversibility plays into it. We develop an abstract conceptual model that we use when discussing the later assumptions.

In section \ref{sec:dynamical_systems} we show how the physical concept of physical distinguishability leads to the mathematical concept of topological spaces. Deterministic and reversible evolution is an invertible continuous map (i.e. a homeomorphism) as it needs to preserve what is physically distinguishable.

In section \ref{sec:reducibility} we introduce the notion of a classical material, one for which we assume infinitesimal reducibility. The state of an arbitrary amount of material is a distribution over the states of the infinitesimal parts, the classical particles. As such a distribution must be invariant under the arbitrary choice of units, we see that for each variable $q^i$ that defines a unit there is a corresponding variable $k^i$ defined on the inverse of that unit. Classical particle states can be identified by conjugate pairs of state variables $(q^i, p_i)$ (i.e. a point in the cotangent bundle $T^*\mathcal{Q}$) and the number of possible initial conditions for each pair is given by $dq^i \wedge dp_i$ (i.e. the canonical symplectic form $\omega$). During evolution, the distribution (and marginal distributions) must be mapped point by point which leads to Hamiltonian mechanics (i.e. a symplectomorphism on $T^*\mathcal{Q}$).

In section \ref{sec:relativity} we show that a relativistic version of Hamilton's equations arises naturally when we consider time dependent laws of evolution for distributions of classical material. This also gives the notion of deterministic and reversible evolution a characterization that is independent of any transformation that mixes time and state variables $(q^i, p_i)$. Moreover it gives us a classical equivalent for anti-particle states.

In section \ref{sec:Lagrangian} we introduce the assumption of kinematic equivalence, studying the motion in space is equivalent to studying the evolution of the state. Under that assumption there is a link between the state variables $(q^i, p_i)$ and the kinematic variables $(x^i, u^i)$ (i.e. position and velocity). This link allows us to express the number of possible initial conditions for a degree of freedom (d.o.f.) in terms of $dx^i$ and $du^i$, leading to the metric tensor $g$. Such an expression also allows us to constrain the Hamiltonian to the one for massive particles under scalar/vector potential forces and an equivalent Lagrangian formulation.

In section \ref{sec:quantum_prelude} we introduce some ideas in classical mechanics to more gently transition to quantum mechanics. We see how distributions over phase space already have features that are typically associated with quantum systems and we will show how to construct a classical uncertainty principle that helps to build intuition used in the later section.

In section \ref{sec:irreducibility} we introduce the notion of an irreducible material, one for which studying the whole distribution does not tell us anything about the motion of its infinitesimal parts, the fragments. A unit amount of this material will correspond to a quantum particle. As this means that the states of such a material are invariant under permutations of the configurations of the fragments, we show that the state space of an irreducible material is a complex vector space. These states can still be compared to each other in a way that gives the state space an inner product. Such comparisons will be preserved by deterministic and reversible evolution which leads to Schroedinger's equation (i.e. a unitary transformation).

In section \ref{sec:discussion} we present a few higher level remarks about the work.

@@

We claim that $\mathsf{T}$ is second countable. Consider the set of outcomes of a process. As each can provide only a finite amount of information (e.g. is identified by a finite number of digits), the set of outcomes is countable. This remains true when combining processes: the number of processes we can combine is finite as is the number of times we can prepare the same state and the set of intersections obtained from the countable outcomes of a finite number of processes is still countable. Consider a basis for the topology. Suppose it is of cardinality greater than countable. Then the basis would be able to distinguish among states in a way that no physical process ever could. The topology could distinguish among states that are not physically distinguishable. The basis for the topology cannot be of cardinality greater than countable. By contradiction, $\mathcal{S}$ must be second countable.
\end{justification}

@@

	We claim that for any fragment variable $f$ there exists a corresponding self-adjoint linear operator $F_{[q^i]} : \Psi(\mathcal{Q}, dq^i) \rightarrow C(\mathcal{Q}, \mathbb{C})$ such that $\int \psi^\dagger(q^i) F_{[q^i]}( \psi(q^i)) d^nq$ is, if it exists, the expectation of $f$ multiplied by the amount of material. Let $f$ be a fragment variable. There exists a corresponding self-adjoint linear operator $F_{[\hat{q}^i]} : \Psi(\hat{\mathcal{Q}}, d\hat{q}^i) \rightarrow C(\hat{\mathcal{Q}}, \mathbb{C})$ where $\hat{\mathcal{Q}}$ provides a maximal set of fragment variables. As both $\hat{\mathcal{Q}}$ and $\mathcal{Q}$ provide a maximal set, we can define $\Upsilon^{q^i}_{\hat{q}^i} \equiv \Upsilon_{\hat{q}^i} \circ \Upsilon_{q^i}^{-1}$ which transforms a wave function over $\mathcal{Q}$ into a wave function over $\hat{\mathcal{Q}}$. $\Upsilon^{q^i}_{\hat{q}^i}$ is an isomorphism as $\Upsilon_{\hat{q}^i}$ and $\Upsilon_{q^i}$ are isomorphisms. $\Upsilon^{q^i}_{\hat{q}^i}$ is unitary as $\Upsilon^{\hat{q}^i}_{q^i} \circ \Upsilon^{q^i}_{\hat{q}^i} = (\Upsilon^{q^i}_{\hat{q}^i})^{-1} \circ \Upsilon^{q^i}_{\hat{q}^i} = I$. We can extend $\Upsilon^{q^i}_{\hat{q}^i}$ on all continuous functions and define $F_{[q^i]} \equiv \Upsilon^{\hat{q}^i}_{q^i} \circ F_{[\hat{q}^i]} \circ \Upsilon^{q^i}_{\hat{q}^i}$. $F_{[q^i]}$ is a self-adjoint linear operator. $\int \psi^\dagger(q^i) F_{[q^i]}( \psi(q^i)) d^nq = \int \psi^\dagger(\hat{q}^i) F_{[\hat{q}^i]}( \psi(\hat{q}^i)) d^n \hat{q}$ is, if it exists, the expectation of $f$ multiplied by the amount of material.
\end{justification}

@@

The state space for classical particles is the one that allows us to express invariant distributions: that is a requirement that comes from the whole. The fragments of quantum particles do not even have a well defined description: the whole tells us nothing about their configuration. In retrospect, this should not be surprising. Nobody is going to hand us down the laws of infinitesimal parts: we induce  what we can by manipulating finite systems. Therefore what the whole allows us to infer about its parts is physically fundamental.

The third notable aspect is that the assumptions have more to do with our ability to create a reliable description rather than essential features of the system itself. A system is not deterministic per se: it is a particular choice for its description (i.e. state) that may undergo deterministic evolution in some cases. In the same vein, a material is not homogeneous per se. Or infinitesimally reducible. In this context, the state, its properties and the laws of evolution are the description of the system under our idealization, not some intrinsic feature of the system.

The fourth intriguing aspect is that, when framing classical and quantum mechanics in this way, there is a progression. First we assume a homogeneous infinitesimally reducible material. Then we assume a homogeneous irreducible material. Extrapolating from this progression, the next assumption would be an inhomogeneous irreducible material: where the material is not in constant equilibrium and its particles do not all look the same.

This different starting point may have potential connections to some of the problems we mentioned in the rationales for the assumptions and to other interesting physics questions in general. In the foreseeable future, however, we will work on the more mundane task of extending this work to field theories, for which a more concrete path is available.

@@

We have seen that much of fundamental physics can be derived from few simple physical assumptions. In doing so we obtained a more unified picture of the different branches of fundamental physics centered around the idea of deterministic and reversible evolution of distributions of homogeneous materials. Though the work is extensive we can summarize a few important conclusions with the following points:

@@

\item Deterministic and reversible evolution is a one-to-one map that preserves the nature of the system (e.g. physical distinguishability, state count and decomposability). Therefore it is an isomorphism on the mathematical structure defined by the state space (e.g. homeomorphism, symplectomorphism, unitary transformation), not just a state-to-state bijection (i.e. isomorphism of sets).

\item Classical Hamiltonian mechanics describes the evolution of a homogeneous material whose state is equivalent to the states of all its infinitesimal parts, the classical particles. The state space for classical particles is a symplectic manifold $(T^*\mathcal{Q}, \omega)$ as this allows us to define state-variable-invariant density distributions; deterministic and reversible evolution for such a state space is a symplectomorphism (i.e. a canonical transformation).

@@

\item Quantum (Hamiltonian) mechanics describes the evolution of a homogeneous material for which the state can only be reduced up to finite parts, the quantum particles, whose internal motion is undetermined. The state space for quantum particles is a complex inner product space $(\Psi, \langle \cdot , \cdot \rangle)$; deterministic and reversible evolution for such a state space is a unitary transformation.
\end{itemize}

While we are sure that some mathematical details can be improved, we feel that the overall ``conceptual castle" is solid. The starting physical assumptions are simple, the translation into math is often straightforward and the intuitive ideas that emerge help unveil connections within and across different branches of math and physics. It is the tight fitting of these different pieces into a broader picture that validates the merit of the approach.

While the scope of this work is limited to refounding and reunderstanding physical theories already experimentally established, it would not be unexpected for a better grasp of old theories to lead to insights and ideas for new ones.

@@

\bibitem{Freidin} Freidin, R., Lasnik, H.: Syntax: critical concepts in linguistics. Routledge (2006)
